### 2018-03-09 paper/abstractions.tex
Finish S2

>>>>

This paper presents a purely functional abstraction for build systems that
allows us to express all the above intricacies of build systems and design
complex build systems from simple primitives. The presented abstraction fits in
just two lines of Haskell code, which are explained
in~\S\ref{sec-abstractions}:

\begin{minted}{haskell}
type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
\end{minted}

### 2018-03-09 paper/background.tex
Finish S2

\begin{table}[h]
\smaller
\centering
\begin{tabular}{l||l|c|c|c|c||l|l}
\hline
$\!\!$Build system$\!$&$\!$Dependencies$\!$&$\!$Minimal$\!$&$\!$Self-tracking$\!$&$\!$Cutoff$\!$&$\!$Cloud$\!$&$\!$Metadata$\!$&$\!$Algorithm$\!\!$\\\hline
$\!\!$\Make       $\!$&$\!$Static  & Yes & No  & No  & No  &$\!$Dirty bit       $\!$&$\!$Topological $\!\!$\\
$\!\!$\Excel      $\!$&$\!$Dynamic & No  & Yes & No  & No  &$\!$Dirty bit, chain$\!$&$\!$Chain, defer$\!\!$\\
$\!\!$\Shake      $\!$&$\!$Dynamic & Yes & No  & Yes & No  &$\!$Graph           $\!$&$\!$Recursive   $\!\!$\\
$\!\!$\Bazel      $\!$&$\!$Static  & Yes & No  & Yes & Yes &$\!$Graphs, cache   $\!$&$\!$Topological $\!\!$\\\hline
$\!\!$This paper  $\!$&$\!$Dynamic & Yes & No? & Yes & Yes &$\!$Graphs, cache   $\!$&$\!$Recursive   $\!\!$\\
\hline
\end{tabular}
\vspace{2mm}
\caption{Summary of build system differences.\label{tab-summary}}
\label{tab:use-cases}
\end{table}

@@

\subsection{Keys, values, hashes and store}

Keys are
used to distinguish values. In build systems keys are typically filenames, e.g.
\cmd{main.c}, whereas values are file contents (a C program source code in this
case). In spreadsheets keys are cell names, e.g. \cmd{A1}, and values are
numbers, text, etc. that are typically displayed inside cells. In Haskell code,
we will use type variables \hs{k} and \hs{v} to denote keys and values,
respectively.

A \emph{store} associates keys to values. It is convenient to assume that a store
is total, i.e. it contains a value for every possible key. We therefore also
assume that the type of values is capable of encoding values corresponding to
non-existent keys (missing files or empty cells).

We use a cryptographic \emph{hash function} \hs{hash :: v -> Hash} for
efficient tracking and sharing of build results.

\subsection{Input, intermediate and output values}

Some values must be provided by the user as \emph{input}. For example,
\textsf{src/file.c} can be edited by the user who relies on the build system to
compile it into \textsf{obj/file.o}. Similarly, the user can input \textsf{A1 = 5}
and \textsf{B1 = 9} expecting the spreadsheet to compute their sum in \textsf{C1},
i.e. \textsf{C1 = 14}.

In the above examples, \textsf{obj/file.o} and \textsf{C1} are \emph{output} values.

In some situations we might also need the notion of \emph{intermediate} values,
which are not interesting for the user but are produced in the process of turning
inputs into outputs. For example, the user might only be interested in the
executable \textsf{bin/file.exe} obtained by linking \textsf{obj/file.o} with
standard libraries, in which case \textsf{obj/file.o} can be considered an
intermediate value.

\subsection{Non-deterministic computations}

Build systems and spreadsheets compute output values from input and intermediate
values. In the most typical case, these \emph{computations} are \emph{functions},
such as \textsf{C1 = A1 + B1}, i.e. their result is uniquely determined by the
input values. However, in general they can be \emph{relations}, i.e. have
multiple valid results. A spreadsheet example: \textsf{A2 = A1 + RANDOM(1,6)}.
This computation has six valid results for each input value \textsf{A1}. In
build systems, the object file \textsf{obj/file.o} is sometimes not uniquely
determined by the source \textsf{src/file.c} -- different compiler runs may
produce different valid results.

\begin{itemize}

    \item Some build systems, e.g. \Buck, require all tasks to be
    \emph{deterministic}, i.e. produce exactly the same output when run on the
    same inputs. However, not all tasks are deterministic, and there are build
    systems that support \emph{non-determinism}. A simple example is \Excel's
    \textsf{RANDBETWEEN(low, high)} that returns a random integer in the
    interval \textsf{[low, high]}.

\end{itemize}

This paper presents a purely functional abstraction for build systems that
allows us to express all the above intricacies of build systems and design
complex build systems from simple primitives. The presented abstraction fits in
just two lines of Haskell code, which are explained
in~\S\ref{sec-abstractions}:

\begin{minted}{haskell}
type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
\end{minted}

Below we define basic notions used in build systems and other similar domains,
for example, spreadsheets.

\subsection{Requirements for build systems}

\begin{itemize}
    \item Correctness
    \item Minimality
    \item Support for sharing and skipping intermediate values
\end{itemize}
...

>>>>

\begin{table}[h]
\smaller
\centering
\begin{tabular}{l||l|c|c|c|c||l|l}
\hline
$\!\!$Build system$\!$&$\!$Dependencies$\!$&$\!$Minimal$\!$&$\!$Self-tracking$\!$&$\!$Cutoff$\!$&$\!$Cloud$\!$&$\!$Metadata$\!$&$\!$Algorithm$\!\!$\\\hline
$\!\!$\Make       $\!$&$\!$Static  & Yes & No  & No  & No  &$\!$Dirty bit       $\!$&$\!$Topological $\!\!$\\
$\!\!$\Excel      $\!$&$\!$Dynamic & No  & Yes & No  & No  &$\!$Dirty bit, chain$\!$&$\!$Chain, defer$\!\!$\\
$\!\!$\Shake      $\!$&$\!$Dynamic & Yes & No  & Yes & No  &$\!$Graph           $\!$&$\!$Recursive   $\!\!$\\
$\!\!$\Bazel      $\!$&$\!$Static  & Yes & No  & Yes & Yes &$\!$Graphs, cache   $\!$&$\!$Topological $\!\!$\\\hline
$\!\!$This paper  $\!$&$\!$Dynamic & Yes & No? & Yes & Yes &$\!$Graphs, cache   $\!$&$\!$Recursive   $\!\!$\\
\hline
\end{tabular}
\vspace{2mm}
\caption{Summary of build system differences.\label{tab-summary}}
\vspace{-7mm}
\end{table}

@@

In this section we introduce a common vocabulary for build systems that allows
us to abstract away from specific application domains, such as software
development or spreadsheets. We will use this vocabulary throughout the rest of
the paper.

\emph{Keys} are used to distinguish \emph{values}. In software build systems
keys are typically filenames, e.g. \cmd{main.c}, whereas values are file
contents (a C program source code in this case). In spreadsheets keys are cell
names, e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
displayed inside cells. In Haskell code, we will use type variables \hs{k}
and \hs{v} to denote keys and values, respectively.

We use a cryptographic \emph{hash function} \hs{hash :: v -> Hash} for efficient
tracking and sharing of build results, where \hs{Hash} is an abstract data type
equipped with the equality test.

A \emph{store} associates keys to values. It is convenient to assume that a store
is total, i.e. it contains a value for every possible key. We therefore also
assume that the type of values is capable of encoding values corresponding to
non-existent keys (missing files or empty cells). This, in particular, suggests
that it is possible to \emph{depend on the absence of a value} which is a useful
feature for build systems. In addition to usual \hs{get} and \hs{put} operations,
some stores support the \hs{getHash} operation implemented more efficiently than
by hashing the result of \hs{get}. For example, GVFS (Git Virtual File
System)~\cite{gvfs} stores file hashes and downloads file contents only on demand.

Some values must be provided by the user as \emph{input}. For example,
\cmd{main.c} can be edited by the user who relies on the build system to
compile it into \cmd{main.o} and subsequently \cmd{main.exe}. End build products,
such as \cmd{main.exe}, are \emph{output} values. All other values (in this case
\cmd{main.o}) are \emph{intermediate}; they are not interesting for the user
but are produced in the process of turning inputs into outputs.

### 2018-03-09 paper/engineering.tex
Finish S2

>>>>

\subsection{Non-deterministic computations}

Build systems and spreadsheets compute output values from input and intermediate
values. In the most typical case, these \emph{computations} are \emph{functions},
such as \textsf{C1 = A1 + B1}, i.e. their result is uniquely determined by the
input values. However, in general they can be \emph{relations}, i.e. have
multiple valid results. A spreadsheet example: \textsf{A2 = A1 + RANDOM(1,6)}.
This computation has six valid results for each input value \textsf{A1}. In
build systems, the object file \textsf{obj/file.o} is sometimes not uniquely
determined by the source \textsf{src/file.c} -- different compiler runs may
produce different valid results.

Some build systems, e.g. \Buck, require all tasks to be
\emph{deterministic}, i.e. produce exactly the same output when run on the
same inputs. However, not all tasks are deterministic, and there are build
systems that support \emph{non-determinism}. A simple example is \Excel's
\textsf{RANDBETWEEN(low, high)} that returns a random integer in the
interval \textsf{[low, high]}.
